---
title: "🐶 Dog Breeds Dashboard"
author: "Jeffrey Gonzales, Morgan Hertel, Amy Gill"
output:
  flexdashboard::flex_dashboard:
    orientation: column
    vertical_layout: fill
    theme:
      version: 4
      bootswatch: journal # some options: simplex, journal,cosmo
      primary: "#8C5A3C"   # warm brown (like a leather collar)
      bg: "#FFF7E6"        # light cream background
      fg: "#4A2C2A"        # dark brown text
runtime: shiny
---

```{r setup, include=FALSE}
# Load packages ------------------------------------------------------------
library(flexdashboard)
library(shiny)
library(tidyverse)
library(plotly)
library(DT)
library(viridis)
library(stringr)

# Read in data -------------------------------------------------------------

breed_rank <- readr::read_csv("data/breed_rank.csv")
breed_traits <- readr::read_csv("data/breed_traits.csv")
trait_description <- readr::read_csv("data/trait_description.csv")

# Clean breed names (fix weird non-breaking spaces) ------------------------

breed_traits <- breed_traits %>%
  mutate(Breed = str_replace_all(Breed, "\u00a0", " "))

# Image lookup table from breed_rank.csv -----------------------------------

breed_images <- breed_rank %>%
  dplyr::select(Breed, ColorImageURL = Color_Image, ImageURL = Image)

# Make long version of rank data: one row per breed–year -------------------

rank_long <- breed_rank %>%
  pivot_longer(
    cols = ends_with("Rank"),      # all the 2013 Rank, 2014 Rank, ...
    names_to = "Year",
    values_to = "Rank"
  ) %>%
  mutate(
    Year = readr::parse_number(Year)  # turn "2013 Rank" → 2013
  )

# Long version of traits: one row per breed–trait ---------------------------

trait_long <- breed_traits %>%
  pivot_longer(
    cols = where(is.numeric),   # only the numeric trait columns
    names_to = "Trait",
    values_to = "Score"
  )

# Order traits by their mean score (low → high) and define a shared color palette

trait_summary <- trait_long %>%
  dplyr::group_by(Trait) %>%
  dplyr::summarise(
    avg_score = mean(Score, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  dplyr::arrange(avg_score)

trait_order <- trait_summary %>% dplyr::pull(Trait)

# One color per trait using viridis (color blind-friendly)-------------------

trait_pal <- viridis::viridis(length(trait_order), option = "H")
names(trait_pal) <- trait_order

# For trait descriptions, make a helper table to join later -----------------

trait_desc_clean <- trait_description %>%
  filter(Trait %in% unique(trait_long$Trait)) %>%  # keep only numeric traits
  select(Trait, Trait_1, Trait_5, Description)

# Summary statistics --------------------------------------------------------

total_breeds <- breed_traits %>% 
  distinct(Breed) %>% 
  nrow()

# Most popular breed in 2020 ------------------------------------------------ 

most_popular_2020 <- breed_rank %>%
  filter(`2020 Rank` == 1) %>%
  pull(Breed)

# Least popular breed in 2020 ----------------------------------------------- 

least_popular_2020 <- breed_rank %>%
  filter(!is.na(`2020 Rank`)) %>%
  arrange(desc(`2020 Rank`)) %>%  # highest rank = least popular
  slice(1) %>%
  pull(Breed)

# Calculate biggest rank improvement (rank going DOWN is better) ------------

biggest_improver <- breed_rank %>%
  mutate(rank_change = `2013 Rank` - `2020 Rank`) %>%
  arrange(desc(rank_change)) %>%
  slice(1) %>%
  pull(Breed)

biggest_improvement <- breed_rank %>%
  mutate(rank_change = `2013 Rank` - `2020 Rank`) %>%
  arrange(desc(rank_change)) %>%
  slice(1) %>%
  pull(rank_change)

# Calculate biggest rank decline (rank going UP = less popular) --------------

biggest_decliner <- breed_rank %>%
  mutate(rank_drop = `2020 Rank` - `2013 Rank`) %>%  # positive = worse rank
  arrange(desc(rank_drop)) %>%
  slice(1) %>%
  pull(Breed)

biggest_rank_drop <- breed_rank %>%
  mutate(rank_drop = `2020 Rank` - `2013 Rank`) %>%
  arrange(desc(rank_drop)) %>%
  slice(1) %>%
  pull(rank_drop)

# Correlation matrix for traits -----------------------------------------------

trait_cor <- breed_traits %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

```

📝 OVERVIEW 
=====================================

Column {data-width=45}
-----------------------------------------------------------------------
### 
```{r echo=FALSE}

tags$div(
  style = "
    display: flex;
    align-items: stretch;     # image stretch vertically
    justify-content: stretch; # image stretch horizontally
    height: 100%;
    width: 100%;
    padding: 0;
    margin: 0;
  ",
  tags$img(
    src = "https://i7.glitter-graphics.org/pub/540/540957x9slahm5jk.gif",
    style = "
      width: 100%;
      height: 100%;
      object-fit: cover;  #fill pane
      border-radius: 0;
      box-shadow: none;
      display: block;
    "
  )
)
```

### **🐾 ABOUT THIS DASHBOARD 🐾**

This interactive flexdashboard uses American Kennel Club (AKC) data to explore dog breeds from different angles:

- **Popularity over time (2013–2020)** using yearly AKC ranks (1 = most popular).  
- **Behavior, care, and suitability traits** scored from 1 (low) to 5 (high).
- **Trait definitions** to help interpret the 1–5 scores (1 = low, 5 = high).
- **Relationships between traits**, such as which characteristics tend to go together.

Across the tabs, you can:

**📝 Overview:** Preview AKC popularity ranks (2013–2020) and how rank works (1 = most popular).

**📊 Summary:** See key metrics (total breeds, most/least popular, biggest riser/faller) plus average trait scores and their distributions.

**🏆 Popularity:** Compare popularity trends for selected breeds over time and view their images, including on hover.

**🔍 Breed Traits Explorer:** Inspect a single breed’s trait profile (1–5 scores), with sketch, full-color photo, and trait definitions.

**🔗 Trait Correlation:** Explore a trait correlation heatmap, top correlated trait pairs, and a scatterplot for any two traits with hoverable breed images.


**Overall, this dashboard is designed to help you:**

- Compare how popular different breeds have been over time.  
- Understand each breed’s behavioral and care traits in detail.  
- Discover patterns and relationships among traits across all breeds

Column {data-width=55}
-----------------------------------------------------------------------
```{r}

# Make a small data frame for the preview ------------------------------------

df_preview <- breed_rank %>%
  dplyr::select(
    Breed,
    `2013 Rank`, `2014 Rank`, `2015 Rank`, `2016 Rank`,
    `2017 Rank`, `2018 Rank`, `2019 Rank`, `2020 Rank`
  )

datatable(
  df_preview,
  options = list(
    pageLength = 10,
    lengthMenu = c(10, 25, 50, 100, nrow(df_preview)),
    scrollY = "70vh",               # taller scrolling region for rows
    scrollCollapse = TRUE,          # collapse if fewer rows
    dom = "lfrtip"                  # keep length, filter, table, info, paging
  ),
  caption = "Sample of Dog Breeds with Popularity Rank (lower = more popular)."
) %>%
  DT::formatStyle(
    columns = names(df_preview),
    backgroundColor = "#FFF7E6",   # warm cream background
    color           = "#4A2C2A"    # deep brown text
  ) %>%
  DT::formatStyle(
    "Breed",
    fontWeight = "bold"            
  )

```


📊 SUMMARY
=====================================

Column {data-width=20}
-----------------------------------------------------------------------

```{r valuebox-css, echo = FALSE, results = "asis"}
cat('
<style>
  # Make value, caption, and icon text white in all valueBoxes
  .value-box .value,
  .value-box .caption,
  .value-box .icon {
    color: #FFFFFF !important;
  }

  # Make all valueBoxes share equal height and fill the pane
  .shiny-valuebox .value-box {
    height: 18vh;                 # 5 boxes ≈ full column height
    display: flex;
    flex-direction: column;
    justify-content: center;      # vertically center value + caption + icon
  }

  /* Optional: smaller gap between boxes */
  .shiny-valuebox {
    margin-bottom: 4px;
  }
</style>
')
```


### Total Breeds in Dataset
```{r}
renderValueBox({
  valueBox(
    value = total_breeds,
    icon = "fa-dog",
    color = "#5E3A24",
  )
})
```

### Most Popular Breed (2020)
```{r}
renderValueBox({
  valueBox(
    value = most_popular_2020,
    icon = "fa-trophy",
    color = "#B66A2A"
  )
})
```

### Least Popular Breed (2020)
```{r}
renderValueBox({
  valueBox(
    value = least_popular_2020,
    icon  = "fa-thumbs-down",
    color = "#555B6E"   
  )
})
```

### Biggest Rank Improvement (2013-2020)
```{r}
renderValueBox({
  valueBox(
    value = paste(biggest_improver, "(+", biggest_improvement, "spots)"),
    icon = "fa-arrow-up",
    color = "#3F7D3C"
  )
})
```

### Biggest Rank Drop (2013-2020)
```{r}
renderValueBox({
  valueBox(
    value = paste(biggest_decliner, "(-", biggest_rank_drop, "spots)"),
    icon  = "fa-arrow-down",
    color = "#9B2226"
  )
})
```




Column {data-width=80 .tabset}
-----------------------------------------------------------------------

### 📈 Average Trait Scores
```{r}

renderPlotly({
  trait_avg <- trait_summary %>%
    dplyr::mutate(Trait = factor(Trait, levels = trait_order))

  p <- ggplot(trait_avg, aes(x = avg_score, y = Trait, fill = Trait)) +
    geom_col() +
    scale_fill_manual(values = trait_pal) +      # shared palette
    scale_x_continuous(limits = c(0, 5), breaks = 0:5) +
    labs(
      x = "Average Score",
      y = NULL,
      title = "Average Trait Scores Across All Dog Breeds"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.title      = element_text(
        hjust = 0.5,   # center title
        face  = "bold",
        size = 12
      ),
       axis.text.y = element_text(
        size = 11,      
        face = "bold"
      ),
      axis.text.x = element_text(size = 11),
      axis.title.x = element_text(size = 12, face = "bold")
    )

  ggplotly(p, tooltip = c("Trait", "avg_score"))
})

```


### 📊 Trait Scores Distribution
```{r}

renderPlotly({
  df <- trait_long %>%
    dplyr::mutate(
      # Reverse the global order so facets start with highest-score trait
      Trait = factor(Trait, levels = rev(trait_order))
    )

  p <- ggplot(df, aes(x = Score, fill = Trait)) +
    geom_histogram(
      bins  = 5,
      color = "#4A2C2A",
      alpha = 0.7
    ) +
    facet_wrap(~ Trait, ncol = 3) +
    scale_fill_manual(values = trait_pal) +      # same colors as avg plot
    scale_x_continuous(breaks = 1:5) +
    labs(
      x = "Score",
      y = "Number of Breeds",
      title = "Distribution of Scores for Each Trait"
    ) +
    theme_minimal() +
    theme(
      strip.text      = element_text(size = 10, face = "bold"),
      axis.text       = element_text(size = 10),
      axis.title      = element_text(size = 12, face = "bold"),
      legend.position = "none",
      plot.title      = element_text(hjust = 0.5, face = "bold", size = 14) # center title
    )

  ggplotly(p)
})

```


**🏆 POPULARITY**
=====================================

Column {data-width=340 .sidebar}
-----------------------------------------------------------------------

### **🐩  BREED SELECTION**

```{r}
inputPanel(
  selectInput(
    inputId  = "breed_select",
    label    = "Please select breed(s) to compare:",
    choices  = sort(unique(rank_long$Breed)),
    selected = c("Retrievers (Labrador)", "French Bulldogs"),
    multiple = TRUE
  ),
  actionButton(
    inputId = "reset_breeds",
    label   = "Reset selection"
  )
)

```

```{r echo=FALSE}
observeEvent(input$reset_breeds, {
  updateSelectInput(
    session,
    "breed_select",
    selected = c("Retrievers (Labrador)", "French Bulldogs")  # reset to defaults
    # or use selected = character(0) to clear everything
  )
})
```

### **🐶 SELECTED BREED**

```{r}

renderUI({
  req(input$breed_select)

  # get the breeds currently selected
  selected_breeds <- input$breed_select

  # look up their images from breed_images
  imgs <- breed_images %>%
    dplyr::filter(Breed %in% selected_breeds)

  # if no images found
  if (nrow(imgs) == 0) {
    return(tags$p("No images available yet for these breeds."))
  }

  # create a small image gallery
  tagList(
    lapply(seq_len(nrow(imgs)), function(i) {
      tags$div(
        style = "display: inline-block;
                 margin: 8px;
                 text-align: center;",
        tags$img(
          src   = imgs$ColorImageURL[i],
          style = "max-width: 160px;
                   height: auto;
                   border-radius: 10px;
                   box-shadow: 0 0 10px rgba(0,0,0,0.2);"
        ),
        tags$div(
          imgs$Breed[i],
          style = "margin-top: 4px; font-size: 0.8em; font-weight: 600;"
        )
      )
    })
  )
})
```

Column {data-width=55}
-----------------------------------------------------------------------
### **📈 POPULARITY RANK TRENDS**

```{r}

renderPlotly({
req(input$breed_select)

#Filter for selected breeds and join image info -------------------------------

df <- rank_long %>%
dplyr::filter(Breed %in% input$breed_select) %>%
dplyr::left_join(breed_images, by = "Breed")

if (nrow(df) == 0) {
return(NULL)
}

#Determine an adaptive upper limit for the rank axis --------------------------

max_rank_sel <- max(df$Rank, na.rm = TRUE)
upper_limit <- max(20, max_rank_sel)

#Choose nice breaks (every 5 ranks) -------------------------------------------

breaks_y <- seq(1, upper_limit, by = 5)

p <- ggplot(
df,
aes(
x = Year,
y = Rank,
color = Breed,
key = Breed # pass Breed as key for hover events
)
) +
geom_line(linewidth = 1) +
geom_point(size = 2) +
scale_y_reverse(
limits = c(upper_limit, 1),
breaks = breaks_y
) +
labs(
x = "Year",
y = "AKC Popularity Rank (1 = most popular)",
title = "Popularity Rank of Selected Dog Breeds Over Time",
color = "Breed"
) +
theme_minimal() +
    theme(
      axis.title.x = element_text(size = 12, face = "bold"),
      axis.title.y = element_text(size = 12, face = "bold"),
      axis.text.x  = element_text(size = 10),
      axis.text.y  = element_text(size = 10),
      plot.title   = element_text(
        hjust = 0.5,
        face  = "bold",
        size  = 14
      )
    )

ggplotly(
p,
tooltip = c("Breed", "Year", "Rank"),
source = "pop_plot" # give this plot a source ID
)
})
```

Column {data-width=45}
-----------------------------------------------------------------------
🐶 BREED IMAGE ON HOVER
```{r}
renderUI({
  # Hover event from this specific plot
  d <- plotly::event_data("plotly_hover", source = "pop_plot")

  if (is.null(d) || nrow(d) == 0) {
    return(
      tags$div(
        style = "
          display: flex;
          align-items: center;      /* vertical center */
          justify-content: center;  /* horizontal center */
          height: 100%;
          width: 100%;
          text-align: center;
        ",
        tags$p("Hover over a point in the popularity plot to see the breed image here.")
      )
    )
  }

  # Breed from 'key' we set in aes(key = Breed)
  breed_hovered <- d$key[1]

  img_url <- breed_images %>%
    dplyr::filter(Breed == breed_hovered) %>%
    dplyr::pull(ColorImageURL) %>%
    .[1]

  if (is.null(img_url) || is.na(img_url) || length(img_url) == 0) {
    return(
      tags$div(
        style = "
          display: flex;
          align-items: center;
          justify-content: center;
          height: 100%;
          width: 100%;
          text-align: center;
        ",
        tags$p(paste("No image available yet for", breed_hovered))
      )
    )
  }

  tags$div(
    style = "
      display: flex;
      align-items: center;      /* vertical center */
      justify-content: center;  /* horizontal center */
      height: 100%;
      width: 100%;
      text-align: center;
    ",
    tags$div(
      tags$h4(breed_hovered),
      tags$img(
        src   = img_url,
        style = "
          max-width: 100%;
          height: auto;
          border-radius: 10px;
          box-shadow: 0 0 10px rgba(0,0,0,0.2);
        "
      )
    )
  )
})

```

```{r echo=FALSE}
tags$div(
  style = "
    text-align: center;
    width: 100%;
    padding: 0;
    margin: 0;
  ",
  tags$img(
    src = "https://i.pinimg.com/1200x/57/78/c3/5778c3d047015bd9996fa884b53ea1a9.jpg",
    style = "
      max-width: 100%;
      height: auto;
      border-radius: 0;
      box-shadow: none;
      display: block;
      margin: 0 auto;
    "
  )
)
```


**🔍 BREED TRAITS EXPLORER **
=====================================

Column {data-width=550 .sidebar}
-----------------------------------------------------------------------

### **PLEASE SELECT DOG BREED** 

```{r}

# Separate dropdown for traits panel

selectizeInput(
  inputId = "breed_traits_select",
  label   = "I want to have a:",
  choices = sort(unique(trait_long$Breed)),
  selected = "Retrievers (Labrador)",   # or NULL if you want no default
  options = list(
    placeholder = 'Type or select a breed...',
    maxItems    = 1
  )
)

```

### **Sketch**

```{r}

# show sketch for the selected breed ------------------------------------------

renderUI({
req(input$breed_traits_select)

# look up image URL from breed_images -----------------------------------------

img_url <- breed_images %>%
dplyr::filter(Breed == input$breed_traits_select) %>%
dplyr::pull(ImageURL)

if (length(img_url) == 0 || is.na(img_url)) {
tags$p("No image available yet for this breed.")
} else {
tags$img(
src = img_url,
style = "max-width: 100%; 
height: auto; 
border-radius: 10px; 
box-shadow: 0 0 10px rgba(0,0,0,0.2);"
)
}
})

```

### **Actual Photo**

```{r}

# show colored image for the selected breed -----------------------------------

renderUI({
req(input$breed_traits_select)

# look up image URL from breed_images -----------------------------------------

img_url <- breed_images %>%
dplyr::filter(Breed == input$breed_traits_select) %>%
dplyr::pull(ColorImageURL)

if (length(img_url) == 0 || is.na(img_url)) {
tags$p("No image available yet for this breed.")
} else {
tags$img(
src = img_url,
style = "max-width: 100%; 
height: auto; 
border-radius: 10px; 
box-shadow: 0 0 10px rgba(0,0,0,0.2);"
)
}
})

```

## Column {.tabset}

### **TRAITS SCORES FOR SELECTED BREED**

```{r}

renderPlotly({
  req(input$breed_traits_select)

  df <- trait_long %>%
    dplyr::filter(Breed == input$breed_traits_select) %>%
    dplyr::left_join(trait_desc_clean, by = c("Trait" = "Trait")) %>%
    dplyr::mutate(
      Trait = factor(Trait, levels = trait_order) 
    ) %>%
    dplyr::arrange(Trait)

  p <- ggplot(df, aes(x = Score, y = Trait, fill = Trait)) +
    geom_col() +
    scale_fill_manual(values = trait_pal) +   # same palette as other plots
    guides(fill = "none") +
    scale_x_continuous(limits = c(0, 5), breaks = 0:5) +
    labs(
      x = "SCORE(1 = low, 5 = high)",
      y = NULL, 
      title = paste("Trait Profile for", input$breed_traits_select)
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.y = element_text(size = 12, face = "bold"),
      axis.text.x = element_text(size = 12, face = "bold"),
      axis.title.x  = element_text(size = 12, face = "bold"),
      plot.title  = element_text(
        hjust = 0.5,
        face  = "bold",
        size  = 14
      )
    )

  ggplotly(p, tooltip = c("Trait", "Score"))
})

```

### **TRAITS EXPLANATION**

```{r}

# Show trait descriptions in an interactive table -----------------------------

datatable(
  trait_desc_clean,
  options = list(
    pageLength = 5,          
    lengthMenu = c(5, 10, 14) 
  ),
  caption = "What each trait means (1 vs 5 on the scale)."
)

```


**🔗 TRAIT CORRELATION**
=====================================

Column {data-width=55 .tabset}
-----------------------------------------------------------------------

### 🌡️ Trait Correlation Heatmap
```{r}

renderPlotly({
  # Convert correlation matrix to long format for ggplot ----------------------
  trait_cor_long <- trait_cor %>%
    as.data.frame() %>%
    rownames_to_column("Trait1") %>%
    pivot_longer(
      cols = -Trait1,
      names_to = "Trait2",
      values_to = "Correlation"
    )
  
  p <- ggplot(trait_cor_long, aes(x = Trait1, y = Trait2, fill = Correlation)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = 0,
      limits = c(-1, 1),
      name = "Correlation"
    ) +
    labs(
      x = "",
      y = "",
      title = "Correlation Between Dog Breed Traits"
    ) +
    theme_minimal() +
   theme(
      axis.text.x = element_text(
        angle = 45,
        hjust = 1,
        size  = 12,
        face  = "bold"
      ),
      axis.text.y = element_text(
        hjust = 1,
        size = 12,
        face = "bold"
      ),
      panel.grid = element_blank(),
      # title – centered, bold, larger
      plot.title = element_text(
        hjust = 0.5,
        face  = "bold",
        size  = 14
      )
    )
  
  ggplotly(p, tooltip = c("Trait1", "Trait2", "Correlation"))
})

```

### 📋 Strongest Correlations
```{r}

renderDT({
  trait_cor_long <- trait_cor %>%
    as.data.frame() %>%
    rownames_to_column("Trait1") %>%
    pivot_longer(
      cols = -Trait1,
      names_to = "Trait2",
      values_to = "Correlation"
    ) %>%
    filter(Trait1 != Trait2) %>%  
    mutate(pair = paste(pmin(Trait1, Trait2), pmax(Trait1, Trait2), sep = " - ")) %>%
    distinct(pair, .keep_all = TRUE) %>%
    select(Trait1, Trait2, Correlation) %>%
    arrange(desc(abs(Correlation))) %>%
    head(10)
  
  datatable(
    trait_cor_long,
    options = list(
      pageLength = 10,
      dom = 't'  
    ),
    caption = "Top 10 strongest correlations between traits (positive or negative)"
  ) %>%
    formatRound("Correlation", 3) %>%
    formatStyle(
      "Correlation",
      backgroundColor = styleInterval(
        cuts = c(-0.5, 0, 0.5),
        values = c("#2166AC", "#F7F7F7", "#F7F7F7", "#B2182B")
      )
    )
})
```

Column {data-width=45}
-----------------------------------------------------------------------

```{r breed_image, server=TRUE}

output$breed_image <- renderUI({
  # Hover event from this specific scatterplot --------------------------------
  d <- plotly::event_data("plotly_hover", source = "trait_scatter")

  # If nothing is hovered yet -------------------------------------------------
  if (is.null(d) || nrow(d) == 0) {
    return(
      tags$div(
        style = "
          display: flex;
          align-items: center;
          justify-content: center;
          height: 100%;
          width: 100%;
          text-align: center;
        ",
        tags$p("Hover over a point in the scatterplot to see the breed image here.")
      )
    )
  }

  # Breed name from 'key' we set in aes(key = Breed) --------------------------
  breed_hovered <- d$key[1]

  # Look up image URL for this breed
  img_url <- breed_images %>%
    dplyr::filter(Breed == breed_hovered) %>%
    dplyr::pull(ColorImageURL) %>%
    .[1]

  if (is.null(img_url) || is.na(img_url) || length(img_url) == 0) {
    return(
      tags$div(
        style = "
          display: flex;
          align-items: center;
          justify-content: center;
          height: 100%;
          width: 100%;
          text-align: center;
        ",
        tags$p(paste("No image available yet for", breed_hovered))
      )
    )
  }

  # Image with name centered just below it ------------------------------------
  tags$div(
    style = "
      display: flex;
      flex-direction: column;
      align-items: center;
      justify-content: center;
      height: 100%;
      width: 100%;
      padding: 0;
      margin: 0;
    ",
    tags$img(
      src = img_url,
      style = "
        max-width: 100%;
        max-height: 100%;
        width: auto;
        height: auto;
        object-fit: contain;
        border-radius: 10px;
        box-shadow: 0 0 10px rgba(0,0,0,0.2);
        display: block;
      ",
      alt = breed_hovered
    ),
    tags$div(
      breed_hovered,
      style = "
        margin-top: 6px;
        font-size: 0.9em;
        font-weight: 600;
        text-align: center;
      "
    )
  )
})
```


### 🔍 Select Traits to Explore Correlation {data-height=40}
```{r}

fluidRow(
  column(
    width = 7,
    tags$div(
      
    ),
    selectInput(
      inputId  = "trait_x",
      label    = "Select X-axis trait:",
      choices  = colnames(trait_cor),
      selected = "Affectionate With Family"
    ),
    selectInput(
      inputId  = "trait_y",
      label    = "Select Y-axis trait:",
      choices  = colnames(trait_cor),
      selected = "Good With Young Children"
    )
  ),
  column(
    width = 5,
    tags$div(
      "🐶 Breed Image",
      style = "
        font-size: 1.0em;
        font-weight: 600;
        margin-bottom: 3px;
      "
    ),
    uiOutput("breed_image")
  )
)

```

### 📉 Scatterplot: Selected Traits {data-height=60}
```{r}

renderPlotly({
  req(input$trait_x, input$trait_y)
  
  df <- breed_traits %>%
    dplyr::select(
      Breed, 
      x_trait = dplyr::all_of(input$trait_x), 
      y_trait = dplyr::all_of(input$trait_y)
    )
  
  # Calculate correlation for these two traits --------------------------------
  cor_value <- cor(df$x_trait, df$y_trait, use = "complete.obs")
  
  p <- ggplot(
    df,
    aes(
      x    = x_trait,
      y    = y_trait,
      text = Breed,   # used in tooltip
      key  = Breed    # used for hover image lookup
    )
  ) +
    geom_jitter(
      width  = 0.1,
      height = 0.1,
      alpha  = 0.6,
      color  = "#8C5A3C",
      size   = 2
    ) +
    geom_smooth(
      method = "lm",
      se     = TRUE,
      color  = "#D4A574",
      fill   = "#D4A574",
      alpha  = 0.2
    ) +
    scale_x_continuous(breaks = 1:5) +
    scale_y_continuous(breaks = 1:5) +
    labs(
      x     = input$trait_x,
      y     = input$trait_y,
      title = paste0("Correlation: ", round(cor_value, 3))
    ) +
    theme_minimal() +
    theme(
      axis.title.x = element_text(size = 12, face = "bold"),
      axis.title.y = element_text(size = 12, face = "bold"),
      axis.text.x  = element_text(size = 10),
      axis.text.y  = element_text(size = 10),
      plot.title   = element_text(
        hjust = 0.5,   # center
        face  = "bold",
        size  = 14
      )
    )
  
  ggplotly(
    p,
    tooltip = c("text", "x", "y"),
    source  = "trait_scatter"   # match the source used in event_data()
  )
})

```



